---
title: "EDA"
author: "Joe LaRocca"
date: "2024-10-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r}

weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = case_match(
      id, 
      "USW00094728" ~ "CentralPark_NY", 
      "USW00022534" ~ "Molokai_HI",
      "USS0023B17S" ~ "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
   month = lubridate::floor_date(date, unit = "month")) |>
  select(name, id, everything())


```

## Making Some Plots

#### Histogram of Precipitation

```{r}

weather_df %>%
  ggplot(aes(x = prcp)) + 
  geom_histogram(fill = "dodgerblue", color = "black")

```

When you see the outliers, look at the data frame to find them, and then you can look up why the outliers exist (for example, the two outliers from New York were in the wake of Hurricanes Henry and Ida in 2021).

```{r}

weather_df %>%
  filter(prcp > 1000)

```

#### Scatterplot of Tmin vs Tmax

```{r}

weather_df %>%
  filter(tmax > 20, tmax < 30) %>%
  ggplot(aes(x = tmin, y = tmax, col = name, shape = name)) + 
  geom_point()

```

## Summarizing Variables

#### Total Observations

```{r}

weather_df %>%
  group_by(name) %>%
  summarize(n_obs = n())

```

```{r}

weather_df %>%
  group_by(month) %>%
  summarize(n_obs = n())

```

```{r}

weather_df %>%
  group_by(name, month) %>%
  summarize(n_obs = n())
```

#### 2x2 tables

```{r}

weather_df %>%
  filter(name != "Molokai_HI") %>%
  drop_na(tmax) %>%
  mutate(
    cold = case_when(
      tmax < 5 ~ "cold",
      tmax >= 5 ~ "not_cold"
    )
  ) %>%
  group_by(name, cold) %>%
  summarize(count = n()) 

```

```{r}

weather_df %>%
  filter(name != "Molokai_HI") %>%
  drop_na(tmax) %>%
  mutate(
    cold = case_when(
      tmax < 5 ~ "cold",
      tmax >= 5 ~ "not_cold"
    )
  ) %>%
  janitor::tabyl(name, cold)

```

## General Numeric Summaries

You can make plots with summary statistics -- just watch out for what you're grouping by!

```{r}

weather_df %>%
  group_by(name, month) %>%
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE),
    median_tmin = median(tmin, na.rm = TRUE),
    sd_prcp = sd(prcp, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = month, y = mean_tmax, color = name)) + 
  geom_point() + 
  geom_line() + 
  theme(legend.position = "bottom")


```

```{r}

weather_df %>%
  group_by(name, month) %>%
  summarize(mean_tmax = mean(tmax, na.rm = TRUE)) %>%
   pivot_wider(
    names_from = name,
    values_from = mean_tmax) %>%
  knitr::kable(digits = 1)

```

## Grouped Mutates

#### Min Rank Function

The `mutate()` function remembers that grouping layers exist!

```{r}

weather_df %>%
  group_by(name) %>%
  mutate(mean_tmax = mean(tmax, na.rm = TRUE),
         centered_tmax = tmax - mean_tmax) %>%
  ggplot(aes(x = date, y = centered_tmax, color = name)) + 
  geom_point()

```

The `min_rank` function is ranking the lowest maximum temperatures in each location, since we grouped by name. "1" represents the lowest max temp, "2" the second-lowest, etc. Using `desc()` within the `min_rank` will put out the maximum high temps instead.

```{r}

weather_df %>%
  group_by(name) %>%
  mutate(
    temp_rank = min_rank(tmax)
  ) %>%
  filter(temp_rank <= 3)

```

```{r}

weather_df %>%
  group_by(name) %>%
  filter(min_rank(tmax) <= 3) %>%
  arrange(tmax)
```

#### Lag Function

The `lag()` function reports the value of a variable from the previous row. For example, lagging the max temp reports the max temp from the previous day.

```{r}

weather_df %>%
  group_by(name) %>%
  mutate(
    lagged_temp = lag(tmax),
    temp_change = tmax - lagged_temp
  ) %>%
  filter(min_rank(temp_change) <= 3)

```

```{r}

weather_df %>%
  group_by(name) %>%
  mutate(
    lagged_temp = lag(tmax),
    temp_change = tmax - lagged_temp
  ) %>%
  summarize(
    mean_tmax_change = mean(temp_change, na.rm = TRUE),
    sd_tmax_change = sd(temp_change, na.rm = TRUE)
  )

```

## Extending Knowledge to Pulse Dataset

```{r}

pulse_df = 
  read_sas("data/public_pulse_data.sas7bdat") 

pulse_df %>%
  janitor::clean_names() %>%
  pivot_longer(
    cols = bdi_score_bl:bdi_score_12m,
    names_to = "visit",
    values_to = "bdi_score",
    names_prefix = "bdi_score_"
  ) %>%
  mutate(visit = ifelse(visit == "bl", "00m", visit)) %>%
  group_by(visit) %>%
  summarize(mean_bdi = mean(bdi_score, na.rm = TRUE),
            median_bdi = median(bdi_score, na.rm = TRUE),
            sd_bdi = sd(bdi_score, na.rm = TRUE))

```

```{r}

pulse_df %>%
  janitor::clean_names() %>%
  pivot_longer(
    cols = bdi_score_bl:bdi_score_12m,
    names_to = "visit",
    values_to = "bdi_score",
    names_prefix = "bdi_score_"
  ) %>%
  mutate(visit = ifelse(visit == "bl", "00m", visit)) %>%
  group_by(visit) %>%
  summarize(
    mean_bdi = mean(bdi_score, na.rm = TRUE)
  ) %>%
  knitr::kable(digits = 1)

```

## FAS Litters

```{r}

litters_df = 
  read_csv("data/FAS_litters.csv", na = c("NA", ".", " ")) %>%
  janitor::clean_names() %>%
  separate(
    group, into = c("dose", "tx_day"), sep = 3
  )

pups_df = 
  read_csv("data/FAS_pups.csv", na = c("NA", ".", " ")) %>%
  janitor::clean_names()

fas_df = 
  left_join(pups_df, litters_df, by = "litter_number")

```

Compute a table that we care about.

```{r}

fas_df %>%
  drop_na(dose) %>%
  group_by(dose, tx_day) %>%
  summarize(mean_pivot = mean(pd_pivot, na.rm = TRUE))

```

```{r}

fas_df %>%
  drop_na(dose) %>%
  group_by(dose, tx_day) %>%
  summarize(mean_pivot = mean(pd_pivot, na.rm = TRUE)) %>%
  pivot_wider(
    names_from = tx_day,
    values_from = mean_pivot
  ) %>%
  knitr::kable(digits = 2)
  

```

